---
title: "Effects of Signaling Environment and Diet Conditions on Mate Choice in the Wolf Spider *Schizocosa mccooki*"
author: "Yihang Park"
date: "2023-12-13"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
mccooki <- read.csv("data/mccooki.csv")
library(tidyverse)
library(dplyr)
library(nnet)
library(car)
library(lmtest)
library(ggplot2)
```

# **Part 0**: Basic data wrangling

Since the raw data is super old and the columns are named and written in an inconvenient manner, let's proceed to tidy up.
```{r}
# since both the visual and vibrational signaling environments are coded in a single column in a peculiar manner (+/+, +/-, -/+, -/-), please separate them into individual columns
# create a binary 'copsuccess' column by converting the 'coptime' column, assigning 1 for copulated and 0 for not copulated
mccooki=mccooki %>% 
  mutate(light.condition=pull(mccooki, Cond..Lt.S.) %>% str_sub(start=1, end=1)) %>% 
  mutate(substrate.condition=pull(mccooki, Cond..Lt.S.) %>% str_sub(start=3, end=3)) %>%
  mutate(cop.success=(X1st.cop.Time!="")+0) 

# change weird column name
colnames(mccooki)[colnames(mccooki) == "X1st.Cop.Male"] ="firstcop.male"

# assign categorical variables as a factor
mccooki$firstcop.male <-as.factor (mccooki$firstcop.male)
mccooki$Fcond <-as.factor (mccooki$Fcond)
mccooki$light.condition <-as.factor(mccooki$light.condition)
mccooki$substrate.condition <-as.factor(mccooki$substrate.condition)
mccooki$Cond..Lt.S. <-as.factor (mccooki$Cond..Lt.S.)
```


Since in the original data(mccooki) response variable (the first male copulated with female) is 3 levels factor including NAs(High-diet vs low-diet vs NA), make another dataframe (mccooki2) with response variable with 2-levels factor excluding NAs (High-diet vs Low-diet).
```{r}
# remove the NAs from the response variable to convert it into a two-level factor (LD vs HD)
mccooki2 <- mccooki %>% 
  filter(!firstcop.male=="na") %>% 
  mutate(firstcop.male = factor(firstcop.male))
```


# **Part 1**: Run various statistical tests!


### **1.1**: Did the signaling environment and/or female diet influence whether females copulated during the trial in the simultaneous choice test?
```{r}
# Performing a generalized linear model with all covariates included
glm.copsuccess <- glm(cop.success~ Fcond*light.condition*substrate.condition, family=binomial(link="logit"),data=mccooki2)
summary(glm.copsuccess)
```
none of the predictors showed significance.

### **1.2**: Did the signaling environment and/or female diet influence the female's choice between males on high and low diets for copulation?

Establish multinomial logistic regression models by progressively including different covariates in a hierarchical manner.
```{r results='hide'}

# let's run multinomial logistic regression analysis  where the response variable is 1st copulated males: Low vs High diet
model1 <-multinom(firstcop.male ~ light.condition ,data=mccooki2)
model2 <-multinom(firstcop.male ~ substrate.condition ,data=mccooki2)
model3 <-multinom(firstcop.male ~ light.condition + substrate.condition ,data=mccooki2) 
model4 <-multinom(firstcop.male ~ light.condition * substrate.condition ,data=mccooki2)
model5 <-multinom(firstcop.male ~ Fcond + light.condition * substrate.condition, data=mccooki2)
model6 <-multinom(firstcop.male ~ Fcond * light.condition * substrate.condition ,data=mccooki2)
```

#### **1.2.1**: getting p-values for each predictors

##### **1.2.1.1**: ANOVA conducted on the full model including all covariates.
```{r}
car::Anova(model6)
```
The presence or absence of vibratory signals affected female choice.



##### **1.2.1.2**: Likelihood ratio tests are conducted to manually compare models and obtain p-values for each predictor.
```{r}
# likelihood ratio test to manually compare the fit of models 
lrtest(model1, model3) #significant effects of vibratory signals
lrtest(model2, model3) #non-significant effects of visual signals
lrtest(model3, model4) #non-significant interaction effects of visual/vibratory signals
lrtest(model4, model5) #non-significant effects of female diet
```
The presence or absence of vibratory signals affected female choice.



##### **1.2.1.3**: To determine the model with the best goodness of fit, let's use a stepwise algorithm to select a model based on AIC. We'll start with a full model containing all covariates (model6).
```{r}
AIC.Bestfit <-step(model6)
```
The model that solely included the vibratory signal environment as a predictor demonstrated the best goodness of fit.Let's use this model to run an ANOVA and obtain the p-value of the predictor.
```{r message=FALSE}
car::Anova(AIC.Bestfit)
```
The reduced/best-fit model also suggests that the presence or absence of vibratory signals affected female choice.

# **Part 2**: Plotting
**Figure 1**: Does the condition of the female and the signaling environment influence copulation success or female receptivity?
Plot bar graph with x axis is differing signaling environment and y axis is a proportion of trials with copulation
```{r fig.width=8, fig.height=6}
# Reorder the sequence of categories displayed on the x-axis
mccooki$Cond..Lt.S.<- factor(mccooki$Cond..Lt.S., levels= c("+/+", "+/-", "-/+", "-/-"))

# To convert the y-axis into a proportion of trials with copulation (ranging from 0 to 1), create a new column indicating the mean copulation success
mccooki <- mccooki %>% 
  group_by(Cond..Lt.S., Fcond) %>% 
  mutate(Proportion = mean(cop.success))

# Plot a bar graph
ggplot(data=mccooki, aes(x = Cond..Lt.S., y = Proportion, fill = Fcond)) +
  geom_bar(stat = "identity", position = "dodge")+
  labs( x= "Signaling environment (Light/Substrate)", y= "Proportions of Trials with Copulation", fill= "Female Diet")+
  scale_fill_discrete(labels=c("h" = "High", "l" = "Low"))+
  ylim(0,1)
```




**Figure 2**: Do female choice vary by signaling environment and male quality?
```{r fig.width=8, fig.height=6}
mccooki2$Cond..Lt.S.<- factor(mccooki2$Cond..Lt.S., levels= c("+/+", "+/-", "-/+", "-/-"))
ggplot(data=mccooki2, aes(x = Cond..Lt.S., y = cop.success, fill = firstcop.male)) +
  geom_col( position = "fill") +
  
# Insert a vertical line at y=0.5, representing the null hypothesis that females exhibit no preference between low and high diet males in this experiment's simultaneous mate choice.
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "black")+
  labs( x= "Signaling environment (Light/Substrate)", y= "Proportions of Trials with Copulation", fill= "Male Diet")+
  scale_fill_discrete(labels=c("h" = "High", "l" = "Low"))
```